---
title: "Consulting Project"
author: "Redacted"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(VGAM)
library(ggplot2)
```

The aim of the project is to first identify whether there is a significant difference in features for different species and to come up with a model that can predict the species based on their features.

# Testing difference in features for different species
## Sepal Length
```{r}
data(iris)
model1 = lm(Sepal.Length ~ Species, data = iris)
summary(model1)
```

## Petal Length
```{r}
model2 = lm(Petal.Length ~ Species, data = iris)
summary(model2)
```

## Petal Width
```{r}
model3 = lm(Petal.Width ~ Species, data = iris)
summary(model3)
```


## Pincipal Component Analysis
As noted from the EDA, the first two principal components capture 95% of the total variability but have significant overlap between versicolor and virginica. We test whether we can come up with a summary statistic of the original data that statistically differ between species.

Using principal component analysis (PCA), we end up with a linear combination of the original variables that capture the most variances among them. The first principal component is approximately proportional to 
$$
Sepal Length + Petal Length + Petal Width - \frac{1}{2}Sepal Width
$$

A similar ANOVA test shows that one can distinguish each species based on this first principal component significantly.

```{r}
pca = princomp(iris[,1:4], cor = T)
pca$loadings[,1]
pca1 = pca$scores[,1]
ggplot(iris, aes(x = Species, y= pca1, col = Species)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  theme(legend.position = "none")
```
```{r}
model_pca = lm(pca1 ~ Species, data = iris)
summary(model_pca)
```

```{r, warning=FALSE}
data(iris)
logistic_model <- vglm(Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, 
                       family = multinomial, data = iris)
predict_probs <- predict(logistic_model, type = 'response')
predict_class <- colnames(predict_probs)[apply(predict_probs, 1, which.max)]
table(iris$Species, predict_class)
```

## Training and Test set

```{r train_test_split, warning=FALSE}
test_index <- seq(1, 150, by = 3)
train <- iris[-test_index, ]
test <- iris[test_index, ]

logistic_train <- vglm(Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, 
                       family = multinomial, data = train)

predict_probs <- predict(logistic_train, newdata = test, type = 'response')
predict_class <- colnames(predict_probs)[apply(predict_probs, 1, which.max)]

table(test$Species, predict_class)
```
